Strong lensing selection effects

### 2022-08-18 paper/seleff.tex
quasar lenses

\def\zsource{z_{\mathrm{s}}}
\def\zlens{z_{\mathrm{g}}}

\def\msource{m_{\mathrm{s}}}

@@

@@

For each of the three intrinsic scatter scenarios, we drew a population of foreground galaxies covering $1000$ square degrees.
The expectation value of the number of galaxies in the corresponding volume, given the redshift and stellar mass cuts described in section \ref{ssub:mstarz}, is YYY.
%Then we generated a population of sources behind the lenses.
%We approximated the foreground galaxies as isolated: for each source, we only modelled the contribution to the lensing field generated by a single lens.
We approximated the foreground galaxies as isolated: when determining whether a galaxy acts as a strong lens, we only modelled the contribution to the lensing signal from the galaxy itself, and neglected that of the environment.
We discuss the possible implications of this approximation in \Sref{sect:discuss}.
For each lens, we determined its caustics relative to the highest source redshift, $\zsource=2.5$. We then placed sources randomly behind the lens. If at least one source fell within a circular region enclosing the caustics, we proceeded to compute its lensed images.

@@

For the simulation with extended sources, we produced images with a pixel size of $0.1''$.
We applied a Moffat PSF with an FWHM of $0.2''$.
Finally, we assumed a background noise level such that an extended source with half-light radius $0.5''$ and an apparent magnitude in the absence of lensing of $\msource=25$ is detected with $S/N=10$.
%This is similar to the background noise expected for the Euclid Wide survey \citep{Sca++22}.
These values are close to those expected for the Euclid Wide survey \citep{Sca++22}.
%
We then applied the peak detection-based lens selection criterion introduced in section \ref{ssec:lensdefext} to find the strong lenses.

%First, we used the criterion described in section \ref{ssec:lensdefext}.
%This resulted in a sample of XXX lenses.
%Then, we added a cut on the Einstein radius of the lens: we only included in the lens sample those with $\tein > 0.5''$.
%This additional cut is meant to approximate the difficulty in detecting strongly lensed images that lie very close to the lens centre: the number of known lenses with Einstein radius smaller than that cutoff value is very small, even in lens surveys that are based on high-resolution data \citep[see e.g.][]{Gar++22}.
%With this more restrictive criterion, the number of strong lenses in the simulated sample is YYY.

%\subsubsection{Lensed quasars}

@@

\Fref{fig:sourcebias} shows the lensing bias in source-related parameters, with respect to the population of sources that are detectable without lensing.
As previously seen in \Fref{fig:sourcepars}, lensed sources are biased towards higher redshift (top panel of \Fref{fig:sourcebias}), with the trend being larger for more restrictive cuts on the Einstein radius.
The second panel shows the bias in the source magnitude. This bias is negative for small values of $\theta_{\mathrm{Ein,min}}$, that is, lensed sources tend to be brighter than their field counterparts.
Naively we would have expected the opposite trend, as strong lensing magnification allows the detection of sources that are intrinsically fainter than the detection limit.
However, as the analysis of \Sref{sect:indlenses} clearly shows, the lensing cross-section is always larger for brighter sources, and that explains the sign of the bias.
At the same time, the median does not capture the whole picture of the bias in $\msource$. For instance, in the fiducial simulation with no cut on the Einstein radius, the $90\%$-ile of the $\msource$ distribution is $26.05$, while that of the population of detectable sources is $0.14$~mag brighter: indeed, strong lensing allows the detection of fainter sources.

>>>>

\def\zsource{z_{\mathrm{s}}}
\def\zlens{z_{\mathrm{g}}}
\def\zqso{z_{\mathrm{qso}}}

\def\msource{m_{\mathrm{s}}}
\def\mqso{m_{\mathrm{qso}}}

@@

We describe the population of background quasars with the following double power-law luminosity function in the rest-frame UV absolute magnitude $M$:
\begin{equation}
\Phi(M,\zqso) = \dfrac{\Phi(M^*)}{10^{0.4(\alpha_Q+1)(M - M^*)} + 10^{0.4(\beta_Q+1)(M - M^*)}}.
\end{equation}
Following \citet{Man++17}, we set $\alpha_Q=-1.35$ (faint-end slope), $\beta_Q=-3.23$ (bright-end slope), and adopted a redshift-evolving normalisation
\begin{equation}
\log{\Phi^*} = -6.0991 + 0.0209\zqso + 0.0171\zqso^2,
\end{equation}
and charachteristic magnitude
\begin{equation}
M^* = -22.5216 - 1.6510\zqso + 0.2869\zqso^2.
\end{equation}
Given the redshift and rest-frame UV luminosity of a quasar, we then computed the apparent magnitude in the observed $i-$band, $\mqso$, using a quasar spectral template\footprint{\url{https://archive.stsci.edu/hlsps/reference-atlases/cdbs/grid/comp_qso/}} built from optical and near-infrared spectra obtained by \citet{vdB++01,Gli++06}.

For the sake of consistency with the population of background extended sources, we limited the redshift distribution of quasars to the range $0.8 < \zqso < 2.5$.
We then truncated the distribution in $\mqso$ at two magnitudes fainter than the detection limit (which is specified in section \ref{ssub:obs}).
Finally, we randomly placed quasars in the source plane with a projected number density of $70$~arcmin$^{-2}$. This is a much larger number density than observed in the real universe. Nevertheless, we are not interested in predicting the absolute number of strong lenses, so this is a viable choice. The advantage of boosting the number density of quasars is that it allows us to produce a large number of lenses without the need for generating too big a population of foreground galaxies.

@@

For each of the three intrinsic scatter scenarios, we drew a population of foreground galaxies covering $1000$ square degrees.
The expectation value of the number of galaxies in the corresponding volume, given the redshift and stellar mass cuts described in section \ref{ssub:mstarz}, is around $300000$.
%Then we generated a population of sources behind the lenses.
%We approximated the foreground galaxies as isolated: for each source, we only modelled the contribution to the lensing field generated by a single lens.
We approximated the foreground galaxies as isolated: when determining whether a galaxy acts as a strong lens, we only modelled the contribution to the lensing signal from the galaxy itself, and neglected that of the environment.
We discuss the possible implications of this approximation in \Sref{sect:discuss}.
For each lens, we determined its caustics relative to the highest source redshift, $\zsource=2.5$. We then placed sources randomly behind the lens. If at least one source fell within a circular region enclosing the caustics, we proceeded to compute its lensed images.

@@

For the simulation with extended sources, we produced images with a pixel size of $0.1''$.
We applied a Moffat PSF with an FWHM of $0.2''$.
Finally, we assumed a background noise level such that an extended source with half-light radius $0.5''$ and an apparent magnitude in the absence of lensing of $\msource=25$ is detected with $S/N=10$.
These values are close to those expected for the Euclid Wide survey \citep{Sca++22}.
%
We then applied the peak detection-based lens selection criterion introduced in section \ref{ssec:lensdefext} to find the strong lenses.

The fiducial scatter simulation with extended background sources produced a sample of $2113$ lenses, corresponding to a number density of $2.1$~deg$^{-2}$. This is about a factor of five smaller than the number density predicted by \citet{Col15} for the Euclid survey. We believe that this is a result of differences in the description of the source population and the criteria used to define a strong lensing event.

@@

\Fref{fig:sourcebias} shows the lensing bias in source-related parameters, with respect to the population of sources that are detectable without lensing.
As previously seen in \Fref{fig:sourcepars}, lensed sources are biased towards higher redshift (top panel of \Fref{fig:sourcebias}), with the trend being larger for more restrictive cuts on the Einstein radius.
The second panel shows the bias in the source magnitude. This bias is negative for small values of $\theta_{\mathrm{Ein,min}}$, that is, lensed sources tend to be brighter than their field counterparts.
Naively one might have expected the opposite trend, as strong lensing magnification allows the detection of sources that are intrinsically fainter than the detection limit.
However, as the analysis of \Sref{sect:indlenses} clearly shows, the lensing cross-section is always larger for brighter sources, and that explains the sign of the bias.
At the same time, the median does not capture the whole picture of the bias in $\msource$. For instance, in the fiducial simulation with no cut on the Einstein radius, the $90\%$-ile of the $\msource$ distribution is $26.05$, while that of the population of detectable sources is $0.14$~mag brighter: indeed, strong lensing allows the detection of fainter sources.
